### 2017-04-02 chapters/introduction/mechanics.tex
Add latest version

In this section we will ...
For a thorough review of continuum mehcanics we refer
to the book of Gerard Holzapel \cite{holzapfel2000nonlinear} from
which most of the theory in this section is taken. For the even more
mathematically oriented reader we refer to \cite{marsden1994mathematical}

\subsection{Introductory continuum mechanics}

\begin{itemize}
  \item Reference to mehcanics book
  \item Introduce basic notation (stress, strain, deformation)
  \end{itemize}

% The mechanical modeling of the heart is based on a continuum approach
% where the heart is represented as a continuum body  embedded in
% $\mathbb{R}^3$.

@@

\subsection{Hyperelasticity}

Even though experimental studies have idicated a viscoelastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider a quasistatic behaviour and therefore
it is also possible to model the myocardium as hyperelastic.
This means that we postulate the existence of
a strain-energy denisity function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$ which is defined as energy per unit reference volume,
and has units $\frac{\text{Joule}}{m^3}$. The first Piola-Kirchhoff
stress for a hyperelastic material is given by th relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}
\end{align}
The strain-energy density function provides us with the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function depends only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been viscoelastic we would expect to see hysterisis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
chage of observer. Formally we must have for any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$ 
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth},
\end{align}
where $\mathrm{Orth}$ is the group of all postive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the matieral is
isotropic, and otherwise we say that the material is anisotropic.

\emph{In stead of introducing the invariants, look at Simones
  thesis. Introduce the invariants in Holzapel and Ogden model}

If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the principle
invariants of $\mathbf{C}$, theat is $\Psi = \Psi(I_1, I_2, I_3)$.
Here $I_i$ are the coefficients in the characteristic polynomial of
$\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversally isotropic
behavior, that is the material has a preferred direction $\mathbf{a}$
we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}\otimes\mathbf{a})\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
denisity function can still be expressed through invariants. However,
we need to included the so called quasi-invariants, which are defined
as streches in the local microstructural coordinate-system. The
transversely isotropic invariants is given by
\begin{align}
  I_4 = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align}
For more detials about invariants see \cite{holzapfel2009constitutive,liu1982representations}.

Recent studies \cite{gultekin2016orthotropic} have raised doubts to
wether the myocardium can be considered purly hyperelastic, and
proposed a viscoelastic model for the myocardium.

\subsection{Incompressibility}

@@

\subsection{Transformations between reference and current
  configuration}
By definition the reference configuration, $\Omega_0$ and current
configuration $\Omega$ are related via the motion $\varphi$ in the
sense that a point $p$ with reference coordinates $\Xvec$ and current
coordinates $\xvec$ satisfies $\xvec = \varphi(\Xvec)$. Likewise a
vector in the $\Omega_0$ is related to a vector in $\Omega$ via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current cofiguration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derrive that the
transformation of an infinitesimal volume element in the reference
cofiguration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current cofiguration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get Nansens formula
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which related normal vector in the current configuration to normal
vector in the reference configuration.

% \subsubsection{Conservation of mass}
% Conservation of mass simply means that for a given motion $\varphi$
% from the reference configuration $\Omega_0$ to the current
% configuration $\Omega$ the mass is conserved. For a material with
% constant density the means that we have
% \begin{align}
%   \int_{\Omega_0} \mathrm{d} V = 
% \end{align}

\subsection{Conservation of linear momentum}

\subsection{Boudary Conditions}
\label{sec:mech_boudary}
Choosing the right boundary conditions for the model is essential,
and the choice should mimic what is observed in the real life. To
physiologically constrain the ventricle in a correct way is difficult,
and many different approaches has been proposed.
The boundary condition and the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure,
\begin{align}
  \mathbf{T} &= -p_{\mathrm{lv}}^{\mathrm{endo}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \mathbf{x} \in \lvendo
\end{align}
This condtion has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\mathbf{T} &= -p_{\mathrm{lv}}^{\mathrm{epi}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \mathbf{x} \in \epi.
\end{align}
However, the pressure $p_{\mathrm{lv}}^{\mathrm{epi}}$ is usually set
to zero. Therefore we drop the superscript and simply refere to the
left ventricular pressure and the left ventricular endocardial
pressure, i.e $p_{\mathrm{lv}} = p_{\mathrm{lv}}^{\mathrm{endo}}$.

The heart in enclosed in the fibroserous sac, called the pericardium
and slides within this sac during a heartbeat. A pericardial type
boundary constraint has been proposed in
\cite{nash2000computational}.

Another approach commonly used, is constrain th longitudinal motion of
base. To apply this type of condition, it is easiest if the base if
flat and located at a prescribed location, for example in the $x= 0$ plane.

Using this
constrain alone is not enough since, the ventricle is free to move in
the basal plane.

Different
models of boundary conditons has been proposed in the litterature,
some begin more popular than others. In Figure we maker

One boundary conditions that is
not debatable is the boundary conditions on the endocardium. The

in order to constrain you model appropratley at the boundary
and

Write about different type of coundary conditions used, in partilar
fixing the base, completely, fixing the endoring, fixing the epiring

\subsection{Force-balance equation}
There are many ways to arrive at the force-balance eqations for the
cardiac mechanics problem. One way is consider conservation of
momentum, which arrives at Cauchys momentum equation. Since we
eventually want to solve this problem using the finite element method,
we only have to consider the equations in its weak form. An intuitive
way of doing this is by utilising a fundamental principle in physics
called the \emph{principle of stationary potential energy}, or
\emph{minimum total potential energy principle}, which states that a
physical system is at equillibrium when the total potential energy is
minmized, and any infinitesimal changes from this state should not add
any energy to the system.

We consider the incompressible case

In orther to make use of this principle we first need to add all the
potential energy in the system. Here we separate between internal and
external energy. Internal energy, is energy that is stored within the
material, for instance when you stretch a rubber band you increase its
internal energy. External energy is all

Denote the domain of interest by $\Omega \subset \mathbb{R}^3$, with
boundary $\partial \Omega$. For the case of a left ventrcular domain
we have $\partial \Omega = \lvendo \cup \lvepi \cup \lvbase$ and for a
bi-ventricular domain we include $\rvendo$ in the partition as well.

For a hyperelastic and incompressible materai, the total potential
energy in the system is give by

\begin{align}
  \Pi(\mathbf{u}, p) &= \Pi_{\mathrm{int}}(\mathbf{u},p) + \Pi_{\mathrm{ext}}(\mathbf{u}). \\
  \Pi_{\mathrm{int}}(u,p) &= \int_{\Omega} \left[ p(J(\mathbf{u}) - 1) +  \Psi(\mathbf{F}(\mathbf{u})) \right] \mathrm{d}V\\
  \Pi_{\mathrm{ext}}(\mathbf{u}) &= - \int_{\Omega} \mathbf{B} \cdot \mathbf{u} \mathrm{d} V - \int_{\partial \Omega_N} \mathbf{T} \cdot \mathbf{u} \mathrm{d}S
\end{align}

Here $\mathbf{B}$ represents body forces acting on a volume element in the reference domain,
e.g  gravity, and $\mathbf{T} = \mathbf{P} \mathbf{N}$ represents
first Piola-Kirchoff traction force acting on a surface element in the
referece domain, e.g pressure.
According to the \emph{Principle of stationary potential energy} we
have 
\begin{align}
  D_{\delta \mathbf{u}} \Pi(\mathbf{u}, p) = 0,  && \text{and} && D_{\delta p} \Pi(\mathbf{u}, p) = 0.
  \label{eq:minimum_potential_energy}
\end{align}
Here $\delta \mathbf{u}$ and $\delta p$ are virtual variations in the
space for the displacement and hydrostatic pressure respectively, and
\begin{align}
  D_{\mathbf{v}} \Phi(\mathbf{x}) = \frac{\mathrm{d}}{\mathrm{d}\varepsilon} \Phi(\mathbf{x} + \varepsilon \mathbf{v})\big|_{\varepsilon = 0}
\end{align}
is the directional derivative of $\Phi$ at $\mathbf{x}$ is the
direction $\mathbf{v}$. The operator is also known as the G\^ateaux
operator. The virtual variations $\delta \mathbf{u}$ and $\delta p$
represents an arbitrary direction with infinitesimal magnitude.

We have:
\begin{align}
  0 = D_{\delta p} \Pi(\mathbf{u}, p)
       % = \frac{\mathrm{d}}{\mathrm{d}\varepsilon} \int_{\Omega} \left(p + \varepsilon \delta p\right)(J(\mathbf{u}) - 1) \mathrm{d}V\\
  = \int_{\Omega}  \delta p(J(\mathbf{u}) - 1) \mathrm{d}V
\end{align}
and
\begin{align}
  \begin{split}
  0 = D_{\delta \mathbf{u}} \Pi(\mathbf{u}, p) %\\
  % =& \frac{\mathrm{d}}{\mathrm{d}\varepsilon} \int_{\Omega}  \left[ p(J(\mathbf{u} + \varepsilon \delta \mathbf{u}) - 1)
    % +  \Psi(\overline{\mathbf{F}}(\mathbf{u}+\varepsilon\delta \mathbf{u})) \right] \mathrm{d}V \\
     % &- \int_{\Omega} \mathbf{B} \cdot \delta \mathbf{u} \mathrm{d} V - \int_{\partial \Omega_N} \overline{\mathbf{T}} \cdot \delta \mathbf{u} \mathrm{d}S\\
  =&  \int_{\Omega}  \left[ pJ \mathbf{F}^{-T} + \mathbf{P} \right] : \Grad \delta \mathbf{u} \mathrm{d}V\\ 
  &- \int_{\Omega} \mathbf{B} \cdot \delta \mathbf{u} \mathrm{d} V - \int_{\partial \Omega_N} \overline{\mathbf{T}} \cdot \delta \mathbf{u} \mathrm{d}S
  \end{split}
\end{align}
We have the following boundary condition
\begin{align}
  u_1 &= 0 , \;  \mathbf{x} \in \lvbase \\
  k\mathbf{u} - \nabla \mathbf{u} \cdot  \mathbf{N}  &= \mathbf{0} , \;  \mathbf{x} \in \lvbase \\
  \mathbf{T} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \mathbf{x} \in \lvendo.
\end{align}
This Dirichlet condition represent fixing the base in the longitudinal ($x$-direction)
direction, and is not present in the weak formuation.
The Robin conditon acts a spring term with sitffness $k \geq 0$ that limits the movement of the
base. The cornercase $k \rightarrow \infty$ represents fixing the base
in all directions while $k = 0$ means that the base are free to move
within the basal plane. The Neumann part of this conditon, $
\mathbf{T} = - \nabla \mathbf{u} \cdot  \mathbf{N}$ is a force. The
robin conditon  enforces that this force should increase linearly with
the displacement. Hence, a larger dispacement will impose a greater
force on the boundary. 
% The problem with this is that any 
The third conditon, i.e the Neumann condtion has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. Plugging in the other two conditions we get the following
force balance equation:
\begin{align}
  \begin{split}
    0 
    % & \int_{\Omega}  \left[ pJ \mathbf{F}^{-T} + \mathbf{P} \right] : \nabla_{\mathbf{X}} \delta \mathbf{u} \mathrm{d}V \\
  % &+ \int_{\partial  \Omega_{i, \mathrm{LV endo}}} p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N} \cdot \delta \mathbf{u} \mathrm{d}S\\
    % &- \int_{ \partial \Omega_{i, \mathrm{base}}} \nabla \mathbf{u} \cdot  \mathbf{N} \cdot \delta \mathbf{u} \mathrm{d}S \\
  =&  \int_{\Omega}  \left[ pJ \mathbf{F}^{-T} + \mathbf{P} \right] : \Grad \delta \mathbf{u} \mathrm{d}V \\
  &+ \int_{\partial  \Omega_{i, \mathrm{LV endo}}} p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N} \cdot \delta \mathbf{u} \mathrm{d}S \\
  &- \int_{ \partial \Omega_{i, \mathrm{base}}} k \mathbf{u} \cdot \delta \mathbf{u} \mathrm{d}S
  \end{split}
\end{align}

\subsection{Discretization of the force balance equations}

The above eqations \eqref{} is only possible to solve analytically for
very simplified problems. Therefore we need to employ a numerial
approxation of the solution of the force balance equaiton. Using the
finite element method we often refer to such approximation as a
Garlerking approximation. This is based on approximating the solutions
by basis functions in a finite dimesional subspace of the true
solution. Let $(V, \langle \cdot \rangle_V)$
and $(Q, \langle \cdot \rangle_Q)$ be two suitable Hilbert
spaces for the displacement $\uvec$ and the hydrostatic pressure $p$
respectively. Now we select some finite dimesional subspaces $V_h
\subset V$ and $Q_h \subset Q$, which are spanned by a finite number
of basis functions.

For incompressible problems such as \eqref{}, we cannot choose the
approximation spaces $V_h, Q_h$ at random. A known numerical issue
that arises is locking, which can be seen if the material do not
deform even if forces are applied. To illustrate the problem of
locking , imagine an incomressible sphere exsposed to higher and higer
hydrostatic pressure. In other words, the hydrostatic pressure is not
unique for a given diplacement. The problem is solved by requirig the
finte element approximation to satisfy the discrete inf-sup condition
\cite{arnold1984stable}. There exist several mixed elements that
satistifies the inf-sup condition. The one used in this thesis is the
Taylor-Hood elements \cite{taylor1973numerical}, with

\subsection{Constitutve relations}

\subsubsection{Modeling of the passive myocardium}
\begin{itemize}
  \item Hyperelasticiy, fung type-refer to relevant papers
  \item Incompressibility - Saddle point problem
  \item Finding the unloaded configuration
\end{itemize}

Talk about the Holzapel Odgen Law, the invariants, the othotropic one,
transvarsally isotropic one, the meaning of each term, analytic
version of the stress

>>>>

In this section we will cover the necessary theory of continuum
mehcanics in order to model the mechanics of the heart. The theory of
continuum mechancs is exstensive, and we will not be able to cover
everything. For a complete review of continuum mechanics the reader is
therefore referred the textbook of Gerard Holzapel
\cite{holzapfel2000nonlinear} from which most of the theory in this
section is taken. For the even more mathematically oriented reader we
refer to \cite{marsden1994mathematical}.

@@

Perhaps a word about stress-strain work conjugates?

@@

\subsubsection{Transformations between reference and current
  configuration}
By definition the reference configuration, $\Omega_0$ and current
configuration $\Omega$ are related via the motion $\varphi$ in the
sense that a point $p$ with reference coordinates $\Xvec$ and current
coordinates $\xvec$ satisfies $\xvec = \varphi(\Xvec)$. Likewise a
vector in the $\Omega_0$ is related to a vector in $\Omega$ via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current cofiguration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derrive that the
transformation of an infinitesimal volume element in the reference
cofiguration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current cofiguration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get \emph{Nanson's formula}
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which related normal vector in the current configuration to normal
vector in the reference configuration.

\subsubsection{Conservation of linear momentum}
Newton's seconds law states that the change in linear momentum equals
the net impulse acting on it. For a continuum materal with constant
mass density $\rho$ this implies that
\begin{align}
  \int_{\Omega} \rho \dot{\mathbf{v}} \mathrm{d}v = \mathbf{f},
  && \mathbf{f} = \int_{\partial \Omega} \mathbf{t} \mathrm{d}s
     + \int_{\Omega} \mathbf{b} \mathrm{d}v,
     \label{eq:cons_lin_mom}
\end{align}
where $\mathbf{v}$ is the spatially velocity field, $\mathbf{t}$ is
the traction acting on the boundary, and $\mathbf{b}$ is the body
force. From \emph{Cauchy's stress theorem} we know that there exist a
second order tensor $\sigma$, known as as Cauchy stress tensor that is
related to the traction vector by $\mathbf{t} = \sigma \mathbf{n}$,
where $\mathbf{n}$ is the unit normal in the current configuration.
Using the divergence theorem we get.

\begin{align*}
  \int_{\partial \Omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \Omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\Omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom}. We arrive at
Cauchys momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution for the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) are negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left}, which is why the force balance
eqautions is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}
  \label{eq:force_balance_strong}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference condiguration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:force_balance_strong_ref}
\end{align}
where $\FPK$ is the first Piola-Kirchoff stress tensor.

\subsubsection{Conservation of angluar momentum}
Just like linear momentum, the angular momentum is also a conserved
quantity. We will not go through the derivation of resulting
eqautions, but state that as a consequense, the Cauchy stress tensor is
symmetric
\begin{align}
  \sigma = \sigma^T
\end{align}

\subsection{Hyperelasticity}
\label{sec:hyperelasticity}

Even though experimental studies have idicated a viscoelastic behavior
of the myocardium \cite{dokos2002shear, gultekin2016orthotropic}, a
common assumption is to consider a quasistatic behaviour and therefore
it is also possible to model the myocardium as hyperelastic.
This means that we postulate the existence of
a strain-energy denisity function $\Psi:\mathrm{Lin}^+ \rightarrow
\mathbb{R}^+$ which is defined as energy per unit reference volume,
and has units $\frac{\text{Joule}}{m^3}$. The first Piola-Kirchhoff
stress for a hyperelastic material is given by th relation
\begin{align}
\FPK = \frac{\partial \Psi(\F)}{\partial \F}
\end{align}
The strain-energy density function relates  the amount of
energy that is stored within the material in response to a given
strain. Hence, the stresses in a hyperelastic material with a given
strain-energy density function, depends only on the strain, and not the
path for which the material deforms. On the contrary, if the model had
been viscoelastic we would expect to see hysterisis in the
stress/strain curve, but this is not possible for a hyperelastic
material.

We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
chage of observer. Formally we must have for any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$ 
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth},
\end{align}
where $\mathrm{Orth}$ is the group of all postive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the matieral is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
chioce of coordinate-system. Having to deal with different
coordinate-systems and mappping quantities from one coordinate-system
to another can results in complicated computation, and the chances of
doing errors are increased. Therefore if would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants, and from their
name it shold be clear that they are invariant with respect to choice
of coordinate-system. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principleinvariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversally isotropic
behavior, that is the material has a preferred direction $\mathbf{a}$
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}\otimes\mathbf{a})\mathbf{Q}^T\},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
denisity function can still be expressed through invariants. However,
we need to included the so called quasi-invariants, which are defined
as streches in the local microstructural coordinate-system. The
transversely isotropic invariants is given by
\begin{align}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed if $\F = \lambda \mathbf{a}_0 \otimes
\mathbf{a}_0$, i.e a pure stretch along $\mathbf{a}_0$, then
$I_{4\mathbf{a}_0 } = \lambda^2$.
For more detials about invariants see \cite{holzapfel2009constitutive,liu1982representations}.

\subsection{Incompressibility}
\label{sec:incompressibility}
An incompressible material is a material in which only isochoric
motions are possible. This means that volume of the material do not
change during any deformation and hence we have the contraint
\begin{align}
  J = \det(\F) = 1.
\end{align}
This constraint can be imposed by considering the strain energy
function
\begin{align}
  \Psi = \Psi(\F) + p(J-1),
  \label{eq:incomp_strain_energy}
\end{align}
where $p$ is a scalar which serves as a Lagrange multiplier, but which
can be identified as the hydrostatic pressure. Like the diplacement,
the hydrostatic is unknown and has to be determined from the equilibrium
equations \eqref{eq:force_balance_strong_ref}. If we differentiate
\eqref{eq:incomp_strain_energy} with respect to $\F$ we get the First
Piola-Kirchhoff stress tensor for an incompressible material
\begin{align}
  \FPK = \frac{\partial \Psi(\F)}{\partial \F} + p \F^{-T}.
\end{align}
Likewise the Cauchy stress tensor is given by
\begin{align}
  \Cauchy = \frac{\partial \Psi(\F)}{\partial \F}\F^{T} + p \I.
  \label{eq:cauchy_incomp}
\end{align}

\begin{remark}
  The sign of $p$ is determined by wether you add or subtract the term
  $ p(J-1)$ to the total strain energy in
  \eqref{eq:incomp_strain_energy}. For all practical purposes, it
  does not matter if you add or subtract the term as long as you are
  consitent. 
\end{remark}

\subsection{Boundary Conditions}
\label{sec:mech_boudary}

Choosing the right boundary conditions for the model is essential,
and the choice should mimic what is observed in the real life. To
physiologically constrain the ventricle in a correct way is difficult,
and many different approaches has been proposed.
The boundary condition and the endocardium is typically modeled as a
Neumann boundary condition, representing the endocardial blood
pressure
\begin{align}
  \sigma \mathbf{n} = -p_{\mathrm{lv}} \mathbf{n}, \;  \xvec \in  \lvendo(t).
\end{align}
This condtion has a negative sign because the unit normal
$\mathbf{N}$ is pointing out of the domain, while the pressure is
acting into the domain. 
Note that this conditon is imposed on the current configuration, and
to utilise the lagrangian formulation we can pull back this condition
to the reference configuration to obtain
\begin{align}
  \FPK\mathbf{N} &= -p_{\mathrm{lv}}^{\mathrm{endo}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo
\end{align}
Likewise, it is common to enforce a
Neumann boundary condition on the epicardium,
\begin{align}
\FPK\mathbf{N}  &= -p_{\mathrm{lv}}^{\mathrm{epi}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \epi.
\end{align}
However, the pressure $p_{\mathrm{lv}}^{\mathrm{epi}}$ is usually set
to zero. Therefore we drop the superscript and simply refere to the
left ventricular pressure as the left ventricular endocardial
pressure, i.e $p_{\mathrm{lv}} = p_{\mathrm{lv}}^{\mathrm{endo}}$.

There exist a variety of boundary conditons at the base.
It is common to constrain the longitudinal motion of
base, even though it is observed in cardiac images that the apex tend
to be more fixed than the base. A recent study shows that taking into
account the base movement is important to capture the correct
geometrical shape \cite{palit2016passive}. However, this has not been
done in the studies in this thesis.
Fixing the longitudinal motion at the base is enforces through a
Dirichlet boundary conditon
\begin{align}
  u_1 = 0,  \;  \Xvec \in \lvbase
\end{align}
To apply this type of condition, it is easiest if the base is
flat and located at a prescribed location, for example in the $x= 0$
plane.
Constraining the longitudinal motion of the base alone is not enough,
since the ventricle is free to move in the basal plane. In order to
anchor the geometry it is possible to fix the movement of the base in
all directions
\begin{align}
  \uvec = \mathbf{0},  \;  \Xvec \in \lvbase,
\end{align}
or fixing the endocardial or epicardial ring
\begin{align}
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{endo}} \\
  \uvec &= \mathbf{0},  \;  \Xvec \in \Gamma_{\mathrm{epi}}.
\end{align}
Another approach which is used in this thesis is to impose a Robin
type boundary conditon
\begin{align}
  \FPK \Nvec + k \uvec = \mathbf{0},  \;  \Xvec \in \lvbase.
\end{align}
Here $k$ can be seen as the stiffness of a spring that limit the
movement at the base. The limiting caes $k = 0$ and $k \rightarrow
\infty$ represent free and fixed base respectively.
Other types of boundary conditions including the pericardium is also
possible \cite{fritz2014simulation} but not considered in this thesis.

\subsection{Force-balance equation}
We will now collect all the terms that makes up to force balance for
the cardiac mechanics problem. Condering the myocardium as a incompressible,
hyperelastic material with a Robin type boundary condition at the base
we obtain the following set of eqations in the Largragian formaulation
\begin{align}
  \begin{split}
  \nabla \cdot \FPK &= 0 \\
  J - 1 &= 0,
  \end{split}
 \label{eq:force_balance_strong}
\end{align}
with $J = \det(\F)$, and the boundary conditions
\begin{align}
  \FPK \Nvec + k \uvec &= \mathbf{0},  \;  \Xvec \in \lvbase\\
  \FPK\mathbf{N}  &= \mathbf{0}, \;  \Xvec \in \epi \\
  \FPK\mathbf{N} &= -p_{\mathrm{lv}} J\mathbf{F}^{-T} \cdot \mathbf{N}, \;  \Xvec \in \lvendo \\
  \label{eq:bndry_conditions_strong}
\end{align}

\subsubsection{Variational formulation}

There are many ways to arrive at the variational formulation of the
force-balance eqations for the cardiac mechanics problem.  One way is
to consider the strong form in \eqref{eq:force_balance_strong} and use
the standard approach in the finite element method to mulitiply by
testfunction in a suitable space, and perform integration by
parts. Within the fields of continuum mechnanics it is common to
forumulate this as a the \emph{principle of virtual work}, with states
that the virtual work of all forces applied to a mechanical system
vanishes in equilibrium. Within this framework, testfunctions are
referred to as virtual variations.
Another approach, which we will use here, derrives the variational form
by utilising a fundamental principle in physics
called the \emph{principle of stationary potential energy}, or
\emph{minimum total potential energy principle}. This principle states that a
physical system is at equillibrium when the total potential energy is
minmized, and any infinitesimal changes from this state should not add
any energy to the system.
In order to make use of this principle we first need to add all the
potential energy in the system. Here we separate between internal and
external energy. Internal energy, is energy that is stored within the
material, for instance when you stretch a rubber band you increase its
internal energy. External energy represent the contribution from all
external forces such as gravity and traction forces.
% We consider the equilibrium in the referece domain, and denote the
% domain of interest by $\Omega \subset \mathbb{R}^3$, with boundary
% $\partial \Omega$. For the case of a left ventrcular domain we have
% $\partial \Omega = \lvendo \cup \lvepi \cup \lvbase$ and for a
% bi-ventricular domain we include $\rvendo$ in the partition as well. 
For a hyperelastic and incompressible material, the total potential
energy in the system is given by
\begin{align}
  \Pi(\mathbf{u}, p) &= \Pi_{\mathrm{int}}(\mathbf{u},p) + \Pi_{\mathrm{ext}}(\mathbf{u}). \\
  \Pi_{\mathrm{int}}(u,p) &= \int_{\Omega} \left[ p(J(\mathbf{u}) - 1) +  \Psi(\mathbf{F}(\mathbf{u})) \right] \mathrm{d}V\\
  \Pi_{\mathrm{ext}}(\mathbf{u}) &= - \int_{\Omega} \mathbf{B} \cdot \mathbf{u} \mathrm{d} V - \int_{\partial \Omega_N} \mathbf{T} \cdot \mathbf{u} \mathrm{d}S
\end{align}
Here $\mathbf{B}$ represents body forces acting on a volume element in
the reference domain, e.g  gravity, and $\mathbf{T} = \mathbf{P}
\mathbf{N}$ represents first Piola-Kirchoff traction force acting on
the Neumann boundary $\partial \Omega_N$. According to the
\emph{Principle of stationary potential energy} we have  
\begin{align}
  D_{\delta \mathbf{u}} \Pi(\mathbf{u}, p) = 0,  && \text{and} && D_{\delta p} \Pi(\mathbf{u}, p) = 0.
  \label{eq:minimum_potential_energy}
\end{align}
Here $\delta \mathbf{u}$ and $\delta p$ are virtual variations in the
space for the displacement and hydrostatic pressure respectively, and
\begin{align}
  D_{\mathbf{v}} \Phi(\mathbf{x}) = \frac{\mathrm{d}}{\mathrm{d}\varepsilon} \Phi(\mathbf{x} + \varepsilon \mathbf{v})\big|_{\varepsilon = 0}
\end{align}
is the directional derivative of $\Phi$ at $\mathbf{x}$ is the
direction $\mathbf{v}$. The operator is also known as the G\^ateaux
operator. The virtual variations $\delta \mathbf{u}$ and $\delta p$
represents an arbitrary direction with infinitesimal magnitude. We have
\begin{align}
  0 = D_{\delta p} \Pi(\mathbf{u}, p)
  = \int_{\Omega}  \delta p(J(\mathbf{u}) - 1) \mathrm{d}V,
\end{align}
and
\begin{align*}
  \begin{split}
  0 = D_{\delta \mathbf{u}} \Pi(\mathbf{u}, p) 
  =&  \int_{\Omega}  \left[ pJ \mathbf{F}^{-T} + \mathbf{P} \right] : \Grad \delta \mathbf{u} \mathrm{d}V\\ 
